Notes on QM in generalized coordinates

### 2023-02-03 Notes/2023-QuantumGeneralizedCoordinates/QuantumGeneralizedCoordinates.tex
Making explicit the derivation of covariance of momentum
Updated from discussion with Lorenzo

\section{Change of coordinates in manifolds}
Transformation rules for differentials on a generic manifold. Only differentiable structure is required.
\begin{equation}
	\begin{aligned}
		\hat{x}^i &= \hat{x}^i(x^j) \\
		d\hat{x}^i &= \partial_j \hat{x}^i dx^j \\
		d\hat{x}^n &= | \partial_j \hat{x}^i | dx^n = |J| dx^n \\
	\end{aligned}
\end{equation}

Transformation rules for a density $\rho(x^i)$. The requirement is that the integral is invariant.
\begin{equation}
	\begin{aligned}
		\int \rho(x^j) dx^n &= \int | \partial_j \hat{x}^i | \hat{\rho}(\hat{x}^i) dx^n = \int \hat{\rho}(\hat{x}^i) d\hat{x}^n \\
		|J| \hat{\rho}(\hat{x}^i) &= | \partial_j \hat{x}^i | \hat{\rho}(\hat{x}^i)  = \rho(x^j) \\
	\end{aligned}
\end{equation}

@@

@@

These are the change of coordinates for momentum. They are derived by making an arbitrary change on the position coordinates and inducing a coordinate change on momentum such that the symplectic form $\omega = dx^i \wedge dp_i$ is invariant. Therefore this depends on the symplectic structure. This makes conjugate momentum vary as a covector.

\begin{equation}
	\begin{aligned}
		\hat{p}_i &= \partial_i x^j p_j \\
		d \hat{p}_i &= \partial_i x^j d p_j \\
		d\hat{p}^n &= | \partial_i x^i | dp^n = |J^{-1}| dx^n \\
	\end{aligned}
\end{equation}

@@

Definition of Fourier transform
\begin{equation}
	\begin{aligned}
		\phi(p_j) &= F(\psi(x^i)) = \frac{1}{\sqrt{2 \pi \hbar}} \int e^{\frac{p_j x^j}{\imath \hbar}} \psi(x^j) dx^n \\
	\end{aligned}
\end{equation}

@@

>>>>

\section{Change of coordinates in manifolds}
Transformation rules for differentials on a generic manifold. Only differentiable structure is required.
\begin{equation}
	\begin{aligned}
		\hat{x}^i &= \hat{x}^i(x^j) \\
		d\hat{x}^i &= \partial_j \hat{x}^i dx^j \\
		d\hat{x}^1 \cdots d\hat{x}^n &= | \partial_j \hat{x}^i | dx^1 \cdots dx^n = |J| dx^1 \cdots dx^n \\
	\end{aligned}
\end{equation}

Transformation rules for a density $\rho(x^i)$. The requirement is that the integral is invariant.
\begin{equation}
	\begin{aligned}
		\int \rho(x^j) dx^1 \cdots dx^n &= \int | \partial_j \hat{x}^i | \hat{\rho}(\hat{x}^i) dx^1 \cdots dx^n = \int \hat{\rho}(\hat{x}^i) d\hat{x}^1 \cdots d\hat{x}^n \\
		|J| \hat{\rho}(\hat{x}^i) &= | \partial_j \hat{x}^i | \hat{\rho}(\hat{x}^i)  = \rho(x^j) \\
	\end{aligned}
\end{equation}

@@

We can rediscover the transformation rules by expanding the state in a different basis.
\begin{equation}
	\begin{aligned}
		\< \psi | \psi \> &= \< \psi | \int dx | x \> \< x | \psi\>  = \< \psi | \int d\hat{x} | \hat{x} \> \< \hat{x} | \psi\> \\
		&= \< \psi | \int ||J|| dx | \hat{x} \> \hat{\psi}(\hat{x}) \\
		&= \< \psi | \int dx \psi(x) \sqrt{||J||} | \hat{x} \>  \\
		| x \> &= \sqrt{||J||} | \hat{x} \>
	\end{aligned}
\end{equation}
Note that $|\psi\> = |\hat{\psi}\>$ because the state is coordinate independent, in the sense that it doesn't change depending in which coordinates is expressed. The delta functions $| x \>$ associated with position, however, do not work in the same way. As these are elements of a continuous spectra, and therefore are associated with densities, we have $| x \>$ $| x \> \neq | \hat{x} \>$ as they depend on the coordinate explicitly (i.e. the normalization constant changes). This would not happen, for example, if the spectra is discrete, like for the eigenstate of a harmonic oscillator.

@@

These are the change of coordinates for momentum. They are derived by making an arbitrary change on the position coordinates and inducing a coordinate change on momentum such that the symplectic form $\omega = dx^i \wedge dp_i$ is invariant. That is:
\begin{equation}
	\begin{aligned}
		\omega &= d\hat{x}^i \wedge d\hat{p}_i = \partial_j \hat{x}^i dx^j \wedge d\hat{p}_i = dx^j \wedge dp_j \\ 
		d p_j &= \partial_j \hat{x}^i d\hat{p}_i
	\end{aligned}
\end{equation}

Therefore this depends on the symplectic structure. This makes conjugate momentum vary as a covector.

\begin{equation}
	\begin{aligned}
		\hat{p}_i &= \partial_i x^j p_j \\
		d \hat{p}_i &= \partial_i x^j d p_j \\
		d\hat{p}^n &= | \partial_i x^j | dp^n = |J^{-1}| dx^1 \cdots dx^n \\
	\end{aligned}
\end{equation}

@@

Definition of Fourier transform
\begin{equation}
	\begin{aligned}
		\phi(p_j) &= F(\psi(x^i)) = \frac{1}{\sqrt{2 \pi \hbar}} \int e^{\frac{p_j x^j}{\imath \hbar}} \psi(x^j) dx^1 \cdots dx^n \\
	\end{aligned}
\end{equation}

@@

\section{Examples}

\subsection{2D Cartesian and 2D polar}

In Cartesian coordinates, the eigenstates of momentum correspond to oscillations at a fixed frequency along the corresponding axis.
\begin{equation}
	\begin{aligned}
		P_x &= - \imath \hbar \partial_x \\
		\phi(x,y) &= e^{\imath k_x x} \\
		P_x e^{\imath k_x x} &= - \imath \hbar \partial_x e^{\imath k_x x} = - \imath \hbar \imath k_x e^{\imath k_x x} = \hbar k_x e^{\imath k_x x} \\ 
		P_y &= - \imath \hbar \partial_y \\
		\phi(x,y) &= e^{\imath k_y y} \\
		P_y e^{\imath k_y y} &= - \imath \hbar \partial_y e^{\imath k_y y} = - \imath \hbar \imath k_y e^{\imath k_y y} = \hbar k_y e^{\imath k_y y} \\ 
	\end{aligned}
\end{equation}

The Fourier transform, then, corresponds to
\begin{equation}
	\begin{aligned}
		\phi(p_x, p_y) &= \frac{1}{\sqrt{2 \pi \hbar}} \int \frac{1}{\sqrt{2 \pi \hbar}} \int e^{\frac{p_x x}{\imath \hbar}} e^{\frac{p_y y}{\imath \hbar}} \psi(x, y) dxdy \\
	\end{aligned}
\end{equation}
where the factors in the transform pick up the oscillations at the appropriate frequency.

In polar coordinates, the eigenstates of momentum in $r$ and $\phi$ are different. For $r$, the oscillation will be along the radial axis (intuitively, like the ripples created by an object falling on water) while for $\phi$ the oscillation will depend on the angle. Eigenstates of momentum for $\phi$ will have to be periodic, and therefore the spectra will be discrete.
\begin{equation}
	\begin{aligned}
		P_r &= - \imath \hbar \partial_r \\
		\phi(r,\phi) &= e^{\imath k_r r} \\
		P_r e^{\imath k_r r} &= - \imath \hbar \partial_r e^{\imath k_r r} = - \imath \hbar \imath k_r e^{\imath k_r r} = \hbar k_r e^{\imath k_r r} \\ 
		P_\phi &= - \imath \hbar \partial_\phi \\
		\phi(r,\phi) &= e^{\imath \frac{ n \phi}{2 \pi}} \\
		P_\phi e^{\imath \frac{ n \phi}{2 \pi}} &= - \imath \hbar \partial_\phi e^{\imath \frac{ n \phi}{2 \pi}} = - \imath \hbar \imath \frac{ n }{2 \pi} e^{\imath \frac{ n \phi}{2 \pi}} = \frac{ \hbar n }{2 \pi} e^{\imath \frac{ n \phi}{2 \pi}r} \\ 
	\end{aligned}
\end{equation}
The Fourier transform in polar coordinates is
\begin{equation}
	\begin{aligned}
		\phi(p_r, p_\phi) &= \frac{1}{\sqrt{2 \pi \hbar}} \int \frac{1}{\sqrt{2 \pi \hbar}} \int e^{\frac{p_r r}{\imath \hbar}} e^{\frac{p_\phi \phi}{\imath \hbar}} \psi(r, \phi) dr d\phi \\
	\end{aligned}
\end{equation}
so that the factors in the transform correctlypick the oscillations on the variables.

Let's double check what happens with coordinate changes.
\begin{equation}
	\begin{aligned}
		x &= r \cos \phi \\
		y &= r \sin \phi \\
		\partial_r x &= \cos \phi \\
		\partial_\phi x &= - r \sin \phi \\
		\partial_r y &= \sin \phi \\
		\partial_\phi y &= r \cos \phi \\
	\end{aligned}
\end{equation}

Express both momentum and eigenstate in other coordinates and see that it works.
\begin{equation}
	\begin{aligned}
		P_r &= \partial_r x P_x + \partial_r y P_y = \cos \phi P_x + \sin \phi P_y = \frac{x}{\sqrt{x^2 + y^2}} P_x + \frac{y}{\sqrt{x^2 + y^2}} P_y  \\
		e^{\imath k_r r} &= e^{\imath k_r \sqrt{x^2 + y^2}}\\
		\partial_x e^{\imath k_r r} &= \partial_x e^{\imath k_r \sqrt{x^2 + y^2}} = e^{\imath k_r \sqrt{x^2 + y^2}} \imath k_r \partial_x \sqrt{x^2 + y^2} = \imath k_r \frac{x}{\sqrt{x^2 + y^2}} e^{\imath k_r \sqrt{x^2 + y^2}} \\
		\partial_y e^{\imath k_r r} &= \partial_y e^{\imath k_r \sqrt{x^2 + y^2}} = e^{\imath k_r \sqrt{x^2 + y^2}} \imath k_r \partial_y \sqrt{x^2 + y^2} = \imath k_r \frac{y}{\sqrt{x^2 + y^2}} e^{\imath k_r \sqrt{x^2 + y^2}} \\
		P_r e^{\imath k_r \sqrt{x^2 + y^2}} &= (-\imath \hbar ) \frac{x^2}{x^2 + y^2} \imath k_r e^{\imath k_r \sqrt{x^2 + y^2}} +(-\imath \hbar ) \frac{y^2}{x^2 + y^2} \imath k_r e^{\imath k_r \sqrt{x^2 + y^2}} \\
		&= \hbar k_r \frac{x^2 + y^2}{x^2 + y^2} e^{\imath k_r \sqrt{x^2 + y^2}} = \hbar k_r e^{\imath k_r \sqrt{x^2 + y^2}} \\
	\end{aligned}
\end{equation}
Note: we haven't transformed the wave function like a density.

Conjecture: does the transform generalize like so:
\begin{equation}
	\begin{aligned}
		\phi(p_{\hat{x}}, p_{\hat{y}}) &= \frac{1}{\sqrt{2 \pi \hbar}} \int \frac{1}{\sqrt{2 \pi \hbar}} \int e^{\frac{p_{\hat{x}} \hat{x}}{\imath \hbar}} e^{\frac{p_{\hat{y}} \hat{y}}{\imath \hbar}} \psi(\hat{x}, \hat{y}) d\hat{x}d\hat{y} \\
	\end{aligned}
\end{equation}
